Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada

### 2021-10-13 analysis/manuscript.Rmd
rolling, rolling, rolling

---
title: Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  html_document:
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document:
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

@@

This is **really fucking bold**.

@@

@@

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $/odmt
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

@@

## Hypothetical plant locations and capacities
For this study, we assumed that forest harvest residue processing would take
place at a plant co-located with an existing forest products manufacturing
facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
pulp mill in Hinton as the locations for the hypothetical forest residue
processing facilities.

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200,000 odmt/yr corresponding to a small wood pellet
mill, 400,000 odmt/yr corresponding to a large pellet mill or a small ethanol
plant, and 800,000 odmt/yr corresponding to a large ethanol plant.

@@

The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone,species, dbh, and height as input variables, so ecozone is important.  It is based on 
equations developed by Lambert
*et al.* (2005) @Lambert2005.

@@

```{r biomass calculator example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
knitr::include_graphics("../img/biomass_calc_example.png")
```

## Spatial variability in residual biomass production

@@

The first step was to identify cells that contained merchantable forest
inventory. We selected cells that contained at least 50 m^3^/ha of merchantable
volume, as that corresponds with merchantability limits specified in Alberta
@Alberta2016. We filtered out cells with more than 50% crown closure of larch
(*Larix spp.*) and birch (*Betula spp.*) as those genera are rarely harvested at
a commercial scale in Alberta.  We filtered out cells with stand age less than
80 years as it is unusual to harvest stands younger than that in Alberta.
This filtered raster was "unioned" with the township coverage, and an average of
the merchantable volume in the merchantable cells was calculated.  This volume
was adjusted upwards by the ratio of the average volume per hectare harvested
from the West Fraser (Hinton) Forest Management Agreement Area (FMA) to the
average merchantable volume for each township calculated from the filtered
raster and township coverages.  The multiplicative volume adjustment factor was
calculated as `r sprintf("%5.2f",merch_vol_adj_factor)`.

@@

Something about needleleaf and broadleaf trees.  Most common species harvested
are white spruce (_Picea glauca_), lodgepole pine (*Pinus contorta* var.
*latifolia*), trembling aspen (*Populus tremuloides*), and less commonly jack
pine (*Pinus banksiana*), black spruce (*Picea mariana*), balsam poplar
(*Populus balsamifera*), subalpine fir (*Abies lasiocarpa*), balsam fir (*Abies
balsamea*), and Douglas-fir (*Pseudotsuga menziesii* var. *glauca*).

## Service areas
** A nice picture demonstrating dendritic nature of road network **
* The provincial road network was sorted into four classes:
* primary provincial highways
* other paved roads
* two-lane gravel roads
* one-lane gravel or dirt roads
* Average speeds calculated using harmonic mean:
$s_h= \left(\frac{s_u^{-1} + s_l^{-1}}{2}\right)^{-1}$
* Speeds by road class, loaded and unloaded:
* primary provincial highways: loaded 80, unloaded 95, h-mean 85
* other paved roads: loaded 65, unloaded 80, h-mean 70
* two-lane gravel roads: loaded 50, unloaded 70, h-mean 60
* one-lane roads: loaded 20, unloaded 30, h-mean 25
* Rasterized network areas were generated using the QGIS plugin, QNEAT3
(Iso-Area as Interpolation) @Raffler2018 .
* QGIS zonal statistics were used to calculate the median network travel time in
each township.

@@

# Theory/Calculation
The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
summaries and statistical analysis were done using R @R2021 using RStudio
@RStudioTeam2021, and tidyverse @Wickham2019.
So here we set the constants.
Following @Roser2013, the pre-piling cost was set to `r
sprintf("%5.2f",prepiling_cost)` \$/odmt, the comminution (*grinding ?*) cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$/odmt, and other costs were
set to  `r sprintf("%5.2f",other_cost)` \$/odmt.  The total "constant" costs are
`r sprintf("%5.2f",constant_cost)` \$/odmt.
Using *blah blah blah* , the truck rate was set to `r
sprintf("%5.2f",truck_rate)` \$/odmt/h.
A 100 m³ trailer can haul `r odt_trailer` odmt of chipped residue.

@@

p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 odmt)") +
  labs(y = "cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,12))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 odmt)") +
  labs(y = "marginal cost ($/odmt)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,175)) 
```

@@

p <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($/odmt)", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))
p

@@

ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($/odmt)") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,150))

@@

Whitecourt is low cost for most years at all capacities.  Noticable spike in
1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
also affects Hinton and Edson at 800,000 odmt capacity, because they reach into the 
Whitecourt area.

>>>>

---
title: Historical Supply of Biomass from Timber Harvest Residue in West-central Alberta, Canada
author:
  - Glen W. Armstrong:
      institute:
        - rr
      correspondence: "yes"
      email: gwa@ualberta.ca
  - Jay A. Anderson:
      email: jaya@ualberta.ca
      institute:
        - rees
  - Alex W. McPhee:
      email: awmcphee@ualberta.ca
      institute: 
        - rees
institute:
  - rr : Department of Renewable Resources, University of Alberta
  - rees: Department of Resource Economics and Environmental Sociology, University of Alberta
date: "`r format(Sys.time(), '%Y-%m-%d')`"
#date: "2021-10-05"
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
output:
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  pdf_document:
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  word_document:
     pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
#csl: canadian-journal-of-forest-research
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

@@

@@

**Do we talk about the weaknesses of BRIMS and how we used 
this to improve?**

@@

#truck_rate <- 6.43 # $/odt/hr
prepiling_cost <- 6.30 # $/odt
comminution_cost <- 25.60 # $/odt
other_cost <- 8.48 # $/odt
constant_cost <- prepiling_cost +  comminution_cost + other_cost # $/Mg DM
location <- c('Whitecourt','Edson','Hinton')
capacity <- c(200000,400000,800000)

@@

## Hypothetical plant locations and capacities
There are a number of companies with Forest Management Agreement Areas
located in or near Edson, Whitecourt, and Hinton.
The West Fraser Mills Ltd. (Hinton) FMA is the primary timbershed for the
Hinton Wood Products sawmill and the Hinton Pulp pulp mill.  The West Fraser
Mills Ltd (Edson) FMA is the primary timbershed for the Edson Forest Products 
sawmill.  The Weyerhaeuser Company Limited (Pembina Timberlands) FMA supplies 
the Edson oriented strand board mill and the Drayton Valley sawmill.  The 
Millar Western FMA supplies sawmills in Whitecourt and Fox Creek, a pulp mill
in Whitecourt, and a specialty wood products facility and Acheson.  The Blue
Ridge Lumber Inc. FMA supplies a sawmill in Blue Ridge, which is about 23 km 
from Whitecourt.  The ANC Timber Ltd. FMA supplies logs to several sawmills in
the area, in exchange for chips used to feed the Alberta Newsprint
Company newsprint mill in Whitecourt.

```{r west_central_ab, echo=FALSE, fig.cap="Location of Forest Management Agreement Areas serving Hinton, Edson, and Whitecourt.", out.width = '80%'}
knitr::include_graphics("../img/west_central_AB_2.png")
```

For this study, we assumed that forest harvest residue processing would take
place at a plant co-located with an existing forest products manufacturing
facility. We used the Weyerhaeuser oriented strand board plant in Edson, the
Millar Western Forest Products Ltd. pulp mill in Whitecourt, and the West Fraser
pulp mill in Hinton as the locations for the hypothetical forest residue
processing facilities.

As well as considering locations for the hypothetical plants, we considered 3
different plant capacities: 200,000 Mg DM/yr corresponding to a small wood pellet
mill, 400,000 Mg DM/yr corresponding to a large pellet mill or a small ethanol
plant, and 800,000 Mg DM/yr corresponding to a large ethanol plant.

@@

The NFIS' individual tree biomass calculator @NFIS2016 uses province, ecozone, 
species, dbh, and height as input variables, so ecozone is important.  It is based on 
equations developed by Lambert
*et al.* (2005) @Lambert2005.

@@

```{r biomass_calculator_example, echo=FALSE, fig.cap="Biomass calculator example.", out.width = '80%'}
knitr::include_graphics("../img/biomass_calc_example.png")
```

```{r taper, echo=FALSE, fig.cap="Volume calcualation example.", out.width = '80%'}
knitr::include_graphics("../img/taper.png")
```
## Spatial variability in residual biomass production

@@

The first step was to identify cells that contained merchantable forest
inventory. We selected cells that contained at least 50 m^3^/ha of merchantable
volume, as that corresponds with merchantability limits specified in Alberta
@Alberta2016. We filtered out cells with more than 50% crown closure of larch
(*Larix spp.* Mill.) and birch (*Betula spp.* L.) as those genera are rarely harvested at
a commercial scale in Alberta.  We filtered out cells with stand age less than
80 years as it is unusual to harvest stands younger than that in Alberta.
This filtered raster was "unioned" with the township coverage, and an average of
the merchantable volume in the merchantable cells was calculated.  This volume
was adjusted upwards by the ratio of the average volume per hectare harvested
from the West Fraser (Hinton) Forest Management Agreement Area (FMA) to the
average merchantable volume for each township calculated from the filtered
raster and township coverages.  The multiplicative volume adjustment factor was
calculated as `r sprintf("%5.2f",merch_vol_adj_factor)`.

@@

Something about needleleaf and broadleaf trees.  Most common species harvested
are white spruce (*Picea glauca* (Moench) Voss), lodgepole pine 
(*Pinus contorta* var. *latifolia* Engelm.), jack pine 
(*Pinus banksiana* Lamb.), trembling aspen (*Populus tremuloides* Michx.),
and less commonly black spruce 
(*Picea mariana* (Mill.) Britton, Sterns & Poggenburg), Engelmann spruce 
(*Picea engelmannii* 
Parry ex Engelm.),  balsam poplar
(*Populus balsamifera* L.), 
balsam fir (*Abies balsamea* (L.) Mill.),  
subalpine fir (*Abies lasiocarpa* (Hooker) Nuttall), 
and Rocky Mountain Douglas-fir 
(*Pseudotsuga menziesii* var. *glauca*  (Mayr) Franco).

## Service areas

The provincial road network was sorted into four classes: primary highways,
other paved roads, two-lane gravel roads, and one-lane gravel or dirt roads.  
Loaded and unloaded travel speeds were assumed for each road class, according
to the table below. The harmonic mean of loaded and unloaded speeds 
(rounded to the nearest
multiple of 5 km/h) were calculated as

$s_h= \left(\frac{s_l^{-1} + s_u^{-1}}{2}\right)^{-1}$.

Use of harmonic mean allows us to calculate the correct round trip travel time.

| Road class      | loaded | unloaded | harmonic mean | 
|:----------------|-------:|---------:|--------------:|
| primary highways|    80  |       95 |           85  |
| other paved     |    65  |       80 |           70  |
| 2-lane gravel   |    50  |       70 |           60  | 
| 1-lane gravel or dirt  |    20  |       30 |           25  | 
| trail           |     ?  |       ?  |           ?   |

Check out https://haozhu233.github.io/kableExtra/awesome_table_in_html.html to
try to format this table better.

* Rasterized network areas were generated using the QGIS plugin, QNEAT3
(Iso-Area as Interpolation) @Raffler2018. **How big are the cells?**

* QGIS zonal statistics were used to calculate the median network travel time in
each township.

@@

# Theory/Calculation
The GIS analysis for this paper was done using QGIS @QGISorg2021.  Data
summaries and statistical analysis were done using R @R2021 using RStudio
@RStudioTeam2021, and tidyverse @Wickham2019.

## Residue yield (Mg DM/ha) by township.

1. Filter out unmerchantable cells from 250 m data
   1. age < 80 years
   2. merchantable volume < 50 m^3^/ha based on 13+/7 cm utilization standard
   8. Bw + Lt prop > 50
2. average merchantable volume of remaining cells to the township level
3. adjust average inventory volume to harvested volume
   - average volume/ha harvested from Hinton FMA / average volume/ha in merchantable inventory from Hinton FMA = 2.38

## Cost calculations

Following @Roser2013, the pre-piling cost was set to
`r sprintf("%5.2f",prepiling_cost)` \$/Mg DM, the comminution (*grinding ?*) cost
was set to `r sprintf("%5.2f",comminution_cost)`  \$/Mg DM, and other costs were
set to  `r sprintf("%5.2f",other_cost)` \$/Mg DM.  The total "constant" costs are
`r sprintf("%5.2f",constant_cost)` \$/Mg DM.
Using *blah blah blah* , the truck rate was set to 
`r sprintf("%5.2f",truck_rate)` \$/Mg DM/h.
A 100 m³ trailer can haul `r odt_trailer` Mg DM of chipped residue.

@@

p <- ggplot(data=cum.odmt, aes(x=residue.cumsum,y=cycleTime,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg DM)") +
  labs(y = "cycle time (hr)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=cycleTime, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000))+
  scale_y_continuous(limits=c(0,12))

@@

```{r more_supply_curves, echo=FALSE, warning=FALSE, fig.cap="Residue supply (delivered cost) curves by location.  Each curve represents one year.  The red curve is for 2009, which was the year of median harvest volume."}
ggplot(data=cum.odmt, aes(x=residue.cumsum,y=residueCost,group=year,colour=as.factor(year))) +
  labs(x = "cumulative residue (1000 Mg DM)") +
  labs(y = "marginal cost ($/Mg DM)") +
  geom_line() +
  scale_colour_manual(values=c("Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Grey","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red","Grey",
                               "Grey","Grey","Grey","Grey","Grey","Red")) +
#  scale_colour_viridis(discrete = TRUE) +
  geom_line(data=cum.odmt[cum.odmt$year==2009,],aes(x=residue.cumsum,y=residueCost, colour="Red"))+
  theme(legend.position = "none") +
  facet_grid(cols=vars(location)) +
  scale_x_continuous(limits=c(0,800000),labels = label_number(scale = 1/1000)) +
  scale_y_continuous(limits=c(50,175)) 
```

@@

p <- ggplot(t,aes(factor(year),residueCost))  + 
  geom_tufteboxplot() +  
  facet_grid(cols=vars(location),rows=vars(capacity)) +
  labs(y = "residue cost ($/Mg DM)", x="year") +
  scale_x_discrete(breaks=c("1990","","","","","1995","","","","","2000","","","","","2005","","","","","2010","","","","","2015")) +
  theme(axis.text.x = element_text(angle=90))
p

@@

ggplot(data=avet, aes(x=year,y=avecost,colour=location))+
  labs(x = "year") +
  labs(y = "average cost ($/Mg DM)") +
  geom_line() +
  facet_grid(rows=vars(capacity)) +
  scale_y_continuous(limits=c(60,150))

@@

Whitecourt is low cost for most years at all capacities.  Noticable spike in
1999.  Probably due to Virginia Hills fire salvage.  Note that this spike
also affects Hinton and Edson at 800,000 Mg DM capacity, because they reach into the 
Whitecourt area.
